Cyclone: Fault Tolerance Middleware for NVM Clusters

### 2016-05-09 paper/paper.tex
Updates to writeup.

\documentclass[twocolumn]{article}
\usepackage{times}

\begin{document}
\title{Cyclone: Fault Tolerance Middleware for NVM Clusters}
\author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
\maketitle
\begin{abstract}
Memory technology in the datacenter is due to undergo a paradigm shift with the
introduction of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as
3D XPoint. Datacenter applications typically use storage devices to survive
faults that take one more machines offline by combining the persistence offered by
durable storage with availability through replication. Building such fault
tolerant applications is hard and users are forced to either build
them from scratch or modify their applications to fit the interfaces of solutions
such as zookeeper. In addition to a loss of flexibility such applications today
are unable to take advantage of the speed of directly attached non-volatile
memory as opposed to older slower storage in the form of disks and SSDs.

@@

\section{Introduction}

\section{Single Node Programming Model}
Cyclone provides users with a replicated and durable heap that can be
manipulated via a set of library routines. The library provides special
functions to allocate and free persistent memory as well as fat pointers that
wrap pointers to persistent memory to ensure that they do not change across
restarts - since the virtual memory ranges into which persistent memory is
mapped may change.

@@

Figure~\ref{fig:example} shows the example of a linked list node and a function
for inserting a new node after a given one. The {\tt TOID} declaration wraps the
linked list node type to produce a fat pointer to it. A {\tt TOID(ll\_node)}
type therefore is a pointer to a linked list node stored in persistent memory.
The {\tt insert\_after} function used the {\tt D\_RO} and {\tt D\_RW} calls to
convert fat pointers into directly accessible read-only and read-write pointers
respectively. Other than the mechanics of using fat pointers the code entirely
resembles that which a programmer might write for a standard linked list in
DRAM.

@@

Cyclone provides tools for programmers to write code that is failure
atomic. Portions of the code are delimited as transactions that are either
completely executed in terms of its effects on persistent memory or not executed
at all. Under the hood, Cyclone tracks all updates to persistent memory in a
persistent undo log - the log entries are persisted before the update to the
actual memory is allowed. On a failure, the undo log is used to roll back
incomplete transactions. A transaction is considered complete when it completes
execution and the updated objects are flushed (automatically by Cyclone) to
persistent memory. The log entries for the transaction are then immediately
deleted. The only burden placed on a programmer for failure atomicity is that
they must inform the Cyclone runtime about objects that are
updated. Figure~\ref{fig:example_fatomic} shows how to convert the function for
adding a node to a linked list into a failure atomic one. The only additions are
the transaction delimiters and the {\tt TX\_ADD} calls to mark the objects being
updated \emph{before} the actual updates to them. Figure~\ref{fig:undo_log}
shows how Cyclone maintains the undo-log of objects.

@@

\end{verbatim}
}
\caption{A Persistent Linked List of Integers}
\label{fig:example_fatomic}
\end{figure}

@@

A crucial feature of Cyclone is that it abstracts the details of the persistence
mechanism away from programmers. This allows the same code to be deployed on
clusters with a persistent storage medium such as 3D XPoint or on clusters with
battery backed NVDIMMs for example. In the former case, explicit {\tt PCOMMIT}
instructions are necessary for persistence [ref PMFS paper], while in the latter
case no such instructions are necessary as on failure the CPU caches and DRAM
are flushed to stable storage under battery (or capacitor) power. Even in the
NVDIMM case however, the undo log can be desirable to protect against OS or
application crashes to ensure that the persistent memory is always left in a
consistent state. Cyclone's programmer visible features are designed to enable
the construction of applications that work regardless of the underlying
persistence mechancisms.

\section{Replication}
Cyclone replicates the state of the NVM heap across multiple machines in a
cluster. Cyclone provides strong consistency - a majority of machines in the
quorum of replicas always agrees on the current state of the NVM heap. Unlike
traditional approaches to strong replication however, Cyclone is designed to
decouple replication from execution of user code modifying NVM. It does so by
encapsulating changes to the NVM heap in RPC calls driven by client requests.
We chose this model since it closely reflects the use cases most suitable for
NVM-based clusters.

Cyclone replicates state \emph{by executing the same sequence of RPC calls on
  all members of the quorum}. By executing the same sequence of RPC calls on all
machines we end up with the same state\footnote{We discuss how to handle
  non-determinism in Section~\ref{sec:synch}}. Cyclone uses RAFT [cite usenix
  raft paper] under the hood for consensus on the contents of the log of RPC
calls - although any other similar state machine replication protocol such as
Paxos is also suitable. Cyclone provides the means to achieve exactly once
semantics to clients by numbering RPC calls and remembering the result of the
last executed RPC call.

\subsection{RPC Call Numbering}
Cyclone is preconfigured for a fixed (large) number of clients. All RPC calls
from a client are numbered and Cyclone remembers the results of the last RPC
call made by a client. A particular client can exist on any machine - cyclone
treats all instances of the client as a single client. Cyclone rejects RPC calls
that are not in sequence. Any RPC call is executed exactly once across the
quorum - regardless of failure and reconfigurations. This provides the means for
clients to easily achieve exactly once semantics, provided the client does not
crash. Cyclone can supply the last executed RPC call number and its result to
assist recovering and stateless clients.

\subsection{Asynchronous Replication}
By default, Cyclone decouples execution of the RPC call on any replica from its
replication. This is particularly beneficial to performance since it allows
overlapping the update of non-volatile state with the network I/O associated
with replication. Persisting RAFT log entries, making the upcall to the
application and executing the RPC call in a failure atomic transaction takes a
significant amount of time and therefore it makes sense to overlap this with the
round-trip delay on the network for replication. This is possible in Cyclone
since we use failure atomic transactions that can be rolled back. Replication
can fail due to various reasons such as failure of nodes, a network partition or
a new leader getting elected in the underlying replication protocol
(RAFT). Since failure atomic transactions can be rolled back, we are able to
speculatively apply the changes to NVM without waiting for replication to
finish.

@@

\begin{figure}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
TX_BEGIN {
  Execute RPC call and modify local NVM
  // Block till replication result known
  if(replication failed) {
    TX_ABORT;
  }
} TX_END
\end{verbatim}
}
\caption{Asynchronous Replication}
\label{fig:async_rep}
\end{figure}

\subsection{Synchronous Replication}
\label{sec:synch}
Asynchronous replication provides good performance but lacks guarantees
occasionally necessary for distributed systems paradigms such as output commit and
the ability to integrate code that has non-deterministic execution paths or side
effects - such as reading the time. Cyclone therefore provides programmers the
ability to closely synchronize execution across the replicas using synchronous
replication. Synchronous replication extends the programming model provided by
cyclone - but neccessitates some awareness on the part of the user about
the difference between the leader and follower replicas. We show later
(Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
fault tolerant applications with little effort on the part of the programmer.

Figure~\ref{fig:sync_rep} shows how synchronous replication operates.  Execution
of the RPC call \emph{waits until the call and its arguments have been
  replicated.} This crucial difference from asynchronous RPC execution ensures
that any externally visible side effects of the RPC such as calls to other
machines only happen after it is guaranteed that the effects of the RPC call
will be committed to the replicas. This provides the output commit property
necessary for some fault tolerance features in applications. Next the RPC call
is executed on the leader and the result of the RPC call is replicated to the
followers. The followers then proceed to execute the RPC call locally taking
into account the result from the leader. This result can include data to
synchronize non-deterministic steps such as reading the time on the replica.

This process is made conditional on the view (current configuration of
replicas and choice of leader) staying unchanged. To successfully complete
execution, the leader must be able to execute the RPC call and replicate the
result within the current view. This ensures that in the event of a network
partition and later repair an old execution of a previous leader cannot commit a
log record with results from a synchronously executed RPC call.

\begin{figure*}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
while(true) {
  TX_BEGIN {
    // Block till replication result known
    if(replication failed) {
      TX_ABORT;
    }  
    if(is_leader && !have_execution_result) {
      Execute RPC call and modify local NVM
      replicate rpc call result in current view;
      // Block till replication result known
      if(result replication failed) {
        TX_ABORT;
      }
    }
    else {
      // Block till replication done or view changes
      if(!result replication complete) {
        TX_ABORT;
      }
      Execute RPC call to modify local NVM with result from leader
    }
  }
} TX_END
\end{verbatim}
}
\caption{Synchronous Replication}
\label{fig:sync_rep}
\end{figure*}

>>>>

\documentclass[twocolumn]{article}
\usepackage{times}
\usepackage{parskip}
\begin{document}
\title{Cyclone: Fault Tolerance Middleware for NVM Clusters}
\author{Amitabha Roy et al. \\ amitabha.roy@intel.com}
\maketitle
\begin{abstract}
Memory technology in the datacenter is due to undergo a paradigm shift with the
introduction of directly addressable non-volatile memory both in the form of
battery backed non-volatile DIMMs as well as newer memory types such as
3D XPoint. Datacenter applications typically use storage devices to survive
faults that take one more machines offline by combining the persistence offered by
durable storage with availability through replication. Building such fault
tolerant applications is hard and users are forced to either build
them from scratch or modify their applications to fit the interfaces of solutions
such as zookeeper. In addition to a loss of flexibility such applications today
are unable to take advantage of the speed of directly attached non-volatile
memory as opposed to older slower storage in the form of disks and SSDs.

@@

%\section{Introduction}

\section{Single Node Programming Model}
Cyclone provides users with a replicated and durable heap that can be
manipulated via a set of library routines. Cyclone builds on top of Intel's Non
Volatile Memory Library (NVML) and provides transparent replication to programs
written to use NVML for persistence on a single node. This section therefore
describes NVML's programming model.

NVML provides special functions to allocate and free persistent memory as well
as fat pointers to persistent memory to ensure that referential integrity in the
heap is maintained across restarts - since the virtual memory ranges into which
persistent memory is mapped may change.

\begin{figure}
{\scriptsize
\begin{verbatim}

@@

Figure~\ref{fig:example} shows the example of a linked list node and a function
for inserting a new node after a given one. The {\tt TOID} declaration wraps the
linked list node type to produce a fat pointer to it. A {\tt TOID(ll\_node)}
type therefore is a pointer to a linked list node stored in persistent memory.
The {\tt insert\_after} function uses the {\tt D\_RO} and {\tt D\_RW} calls to
convert fat pointers into directly accessible read-only and read-write pointers
respectively. Other than the mechanics of using fat pointers the code entirely
resembles that which a programmer might write for a standard linked list in
DRAM.

@@

NVML provides tools for programmers to write code that is failure
atomic by allowing a block of code to be delimited as a transaction. A
transaction is either completely executed in terms of its effects on
persistent memory or not executed at all. Under the hood, NVML tracks all
updates to persistent memory in a persistent undo log - the log entries are
persisted before the update to the actual memory is allowed. On a failure, the
undo log is used to roll back incomplete transactions. A transaction is
considered complete when it completes execution and the updated objects are
flushed (automatically by NVML) to persistent memory. The log entries for the
transaction are then immediately deleted. The only burden placed on a programmer
for failure atomicity is to delimit transactions and that they must inform NVML
about objects that are updated. Figure~\ref{fig:example_fatomic} shows how to
convert the function for adding a node to a linked list into a failure atomic
one. The only additions are the transaction delimiters and the {\tt TX\_ADD}
calls to mark the objects being updated \emph{before} the actual updates to
them. Figure~\ref{fig:undo_log} shows how Cyclone maintains the undo-log of
objects.

@@

\end{verbatim}
}
\caption{Failure Atomic Peristent Linked List}
\label{fig:example_fatomic}
\end{figure}

@@

A crucial feature of NVML is that it abstracts the details of the persistence
mechanism away from programmers. This allows the same code to be deployed on
clusters with a range of persistence mechanisms - mapping pages from secondary
storage such as SSDs, using battery backed NVDIMMs, using an onboard battery to
save DRAM state to SSD on failure [FARM] or using directly accessed persistent
memory [PMFS]. Each of these options has different requirements. For example the
NVM used in PMFS requires explicit cache flush (CLFLUSH) and PCOMMIT
instructions. NVDIMMs would conceivably require only a CLFLUSH to flush the
cacheline and ensure that the line enters the asynchronous data refresh
domain. Using secondary storage is also an option with faster 3DXPoint storage
behind an NVMe interface - this would require an {\tt msync} and OS pagecache
intervention to persist data to secondary storage. NVML seamlessly uses the
right persistence mechanism without requiring code using the library to be aware
of these details.

\section{Replication}
Cyclone replicates the state of the NVML heap across multiple machines in a
cluster. Cyclone provides strong consistency - a majority of machines in the
quorum of replicas always agrees on the current state of the NVM heap. Unlike
traditional approaches to strong replication however, Cyclone is designed to
decouple replication from execution of user code modifying NVM. It does so by
encapsulating changes to the NVM heap in server side code driven by client RPC
requests. We chose this model since it closely reflects the use cases most
suitable for NVM-based clusters.

Cyclone replicates state \emph{by executing the same sequence of RPC calls on
  all members of the quorum}. By executing the same sequence of RPC calls on all
machines we end up with the same state\footnote{We discuss how to handle
  non-determinism in Section~\ref{sec:synch}}. Cyclone uses RAFT [cite usenix
  raft paper] under the hood for consensus on the log of RPC calls - although
any other similar state machine replication protocol such as
Paxos is also suitable. Cyclone provides the means to clients to achieve exactly
once semantics for the RPC calls they make by numbering RPC calls and
remembering the number and result of the last executed RPC call at the server end.

\subsection{RPC Call Numbering}
Cyclone is preconfigured for a fixed (large) number of clients. All RPC calls
from a client are numbered and Cyclone remembers (in a persistent manner) the
number and result of the last RPC call made by a client. A particular client can
exist on any machine - cyclone treats all instances of the client as a single
client. Cyclone rejects RPC calls that are not in sequence. Any RPC call is
executed exactly once across the quorum - regardless of failure and
reconfigurations. This provides the means for clients to easily achieve exactly
once semantics, provided the client does not crash. Cyclone can supply the last
executed RPC call number and its result to assist recovering and stateless
clients.

\subsection{Asynchronous Replication}
By default, Cyclone decouples execution of the RPC call on any replica from its
replication. This is particularly beneficial to performance since it allows
overlapping the update of non-volatile state with the network I/O associated
with replication. Making the call to the application and executing the it in a
failure atomic transaction takes a
significant amount of time and therefore it makes sense to overlap this with the
round-trip delay on the network for replication. This is possible in Cyclone
since we use failure atomic transactions that can be rolled back. Replication
can fail due to various reasons such as failure of nodes, a network partition or
a new leader getting elected in the underlying replication protocol
(we refer to this as a view change). Since failure atomic transactions can be
rolled back, we are able to speculatively apply the changes to NVM without
waiting for replication to finish.

@@

\begin{figure}
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
TX_BEGIN {
  Execute RPC call and modify local NVM
  Block till replication result known
  if(replication failed) {
    TX_ABORT;
  }
} TX_END
\end{verbatim}
}
\caption{Asynchronous Replication}
\label{fig:async_rep}
\end{figure}

\subsection{Synchronous Replication}
\label{sec:synch}
Asynchronous replication provides good performance but lacks guarantees
occasionally necessary for distributed systems paradigms such as output commit
and the ability to integrate code that has non-deterministic execution paths
(such as being dependent on the current time) or side effects - such as making
further requests. Cyclone therefore provides programmers the ability to closely
synchronize execution across the replicas using synchronous
replication. Synchronous replication extends the programming model provided by
cyclone - but neccessitates some awareness on the part of the user about the
difference between the leader and follower replicas. We show later
(Sec~\ref{sec:examples}) how this awareness enables the crafting of efficient
fault tolerant applications with little effort on the part of the programmer.

Figure~\ref{fig:sync_rep} shows how synchronous replication operates.  Execution
of the RPC call \emph{waits until the call and its arguments have been
  replicated.} This crucial difference from asynchronous RPC execution ensures
that any externally visible side effects of the RPC such as calls to other
machines only happen after it is guaranteed that the effects of the RPC call
will be committed to the replicas. This provides the output commit property
necessary in some applications.

Next, the RPC call is executed on the leader and the result of the RPC call is
replicated to the followers. The followers then proceed to execute the RPC call
locally taking into account the result from the leader. This result can include
data to synchronize non-deterministic steps such as reading the time on the
replica.

This process is made conditional on the view (current configuration of
replicas and choice of leader) staying unchanged. To successfully complete
execution, the leader must be able to execute the RPC call and replicate the
result within the current view. This ensures that in the event of a network
partition and later repair, an old execution of a previous leader cannot commit a
log record with results from a synchronously executed RPC call.

\begin{figure*}
\centering
\begin{minipage}{0.7\textwidth}  
{ \scriptsize
\begin{verbatim}
Initiate replication of RPC call and arguments
Block till replication of RPC call returns
if(replication success) {  
  while (true) { // Must complete 
    view = current RAFT view (term)
    TX_BEGIN {
      if(is_leader && !have_execution_result) {
        Execute RPC call and modify local NVM
        replicate result in view;
        Block till result replication returns
        if(result replication failed) {
          TX_ABORT;
        }
      }
      else {
        Block till committed result replication or view changes
        if(!committed result replication) {
          TX_ABORT;
        }
        Execute RPC call to modify local NVM with result from leader
      }
    } TX_ONABORT {
      continue; // Try again
    } TX_END
    break; // All done
  }
}
\end{verbatim}
}
\end{minipage}
\caption{Synchronous Replication}
\label{fig:sync_rep}
\end{figure*}
